---
title: "Getting started with ccdR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{"Getting started with ccdR"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = (nchar(Sys.getenv('CCTE_API_KEY')) > 0)
)
```


```{r setup}
if (!library(ccdR, logical.return = TRUE)){
  devtools::load_all()
}
```

# ccdR quickstart guide

The [CompTox Chemicals Dashboard](https://comptox.epa.gov/dashboard/) (CCD) provides a centralized location for data on chemicals. Such data includes phys-chem properties, hazard details, and bioactivity, among others. For users interested in looking up a few chemicals and using the associated data, the website provides a great resource. Moreover, for those interested in batch searching a list of chemicals, the CCD also supplies this functionality, offering a variety of output formats from which one can choose.

However, the batch search functionality of the CCD does not necessarily fit well into a programmatic approach to handling chemical data. For this, accessing the information available on the CCD through an API is a better solution. And recently, such an API became publicly available; the [CCTE APIs](https://api-ccte.epa.gov/docs/index.html) provide access to this information.

Users who are not familiar or experienced with using APIs may not find it appealing learning and writing code to access the APIs, so ccdR provides a more friendly alternative to this. 


## API key

Before getting started, a user must have an API key ready to use in order to access the CCTE APIs. This can be obtained from the admins of the CCTE APIs by emailing [CCTE API Admins](mailto:ccte_api@epa.gov). In this vignette, the key will be stored as the variable `my_key`.

To run this vignette locally, execute the following code prior to knitting the vignette (in fact, prior to downloading the package). 

```{r, eval=FALSE}
Sys.setenv('CCTE_API_KEY' = '<YOUR API KEY>')
```

Next, we store the api key in the variable `my_key`.

```{r, api-key}
my_key <- Sys.getenv('CCTE_API_key')
```

For general use of the package, one may use the function `register_ccdr()` to store the API key in the current session or more permanently for access across sessions. The following example demonstrates this.

```{r, register-ccte, eval=FALSE}
# This stores the key in the current session
register_ccdr(key = '<YOUR API KEY>')

# This stores the key across multiple sessions and only needs to be run once. If the key changes, rerun this with the new key.
register_ccdr(key = '<YOUR API KEY>', write = TRUE)
```

The API key is stored and the default display setting is turned off for protection. To change this, use the following functions as demonstrated.

```{r, display-hide-key}
# To show the API key
ccdr_show_api_key()
getOption('ccdr')$display_api_key

# To hide the API key
ccdr_hide_api_key()
getOption('ccdr')$display_api_key
```

Finally, to access the key, use the `ccte_key()` function.

```{r, ccte-key, eval = FALSE}
ccte_key()
```



## Chemical APIs

The CCTE APIs are split into three sections, the first of which is [Chemical APIs](https://api-ccte.epa.gov/docs/chemical.html#/). In the next several code chunks, we demonstrate how to use functions from ccdR to access different types of chemical information.

### Chemical Details Resource

The first function we explore is `get_chemical_details()`, which takes in either the DTXSID or DTXCID of a chemical and the user-specific API key. We obtain the relevant chemical details for Bisphenol A (BPA), which has DTXSID "DTXSID7020182".

```{r, bpa-chem-details}
bpa_details <- get_chemical_details(DTXSID = 'DTXSID7020182',
                                    API_key = my_key)
head(bpa_details)
```
The output of this function is a data.table. We could have also input the DTXCID rather than the DTXSID and returned the same data.

### Chemical Property Resource

The function `get_chem_info()` is similar. This returns phys-chem properties, which can be filtered to 'experimental' or 'predicted' if desired.


First, we grab all results. 

```{r, bpa-chem-info}
bpa_info <- get_chem_info(DTXSID = "DTXSID7020182",
                          API_key = my_key)

head(bpa_info)
```

Next, we filter by experimental results.

```{r, bpa-experimental}
bpa_info_experimental <- get_chem_info(DTXSID = "DTXSID7020182",
                                       type = 'experimental',
                                       API_key = my_key)

head(bpa_info_experimental)
```

Finally, we filter by predicted results.

```{r, bpa-predicted}
bpa_info_predicted <- get_chem_info(DTXSID = "DTXSID7020182",
                                    type = 'predicted',
                                    API_key = my_key)

head(bpa_info_predicted)
```

### Chemical Fate Resource

To gather chemical fate details, we supply the same information as we did for *bpa_info* to the function `get_fate_by_dtxsid()`.

```{r, fate}
bpa_fate <- get_fate_by_dtxsid(DTXSID = 'DTXSID7020182',
                               API_key = my_key)
head(bpa_fate)
```


### Chemical Search Resource

The next set of functions operate by searching for chemicals that match an input character string; those that start with, those that contain, and those that equal the input string. Each takes in a string and the API key and each returns a data.frame with 11 columns, giving various details of the matched chemicals.

We demonstrate this on the function `chemical_equal()`, though the behavior of the functions `chemical_contains()` and `chemical_starts_with()` are analogous.

```{r, bpa-equal}
bpa_equals <- chemical_equal(word = 'Bisphenol a',
                             API_key = my_key)
head(bpa_equals)
```
Next we demonstrate the family of ms ready functions. These return lists of DTXSIDs of chemicals that match the search criteria. For the example, we use Methane, with chemical formula CH4, mono-isotopic mass of 16.0313g, and DTXCID of DTXSID8025545.

First, we use the formula for methane and display the first ten responses.

```{r, msready-formula}
methane_formula <- get_msready_by_formula(formula = 'CH4', 
                                          API_key = my_key)
length(methane_formula)
methane_formula[1:10]
```

Next, we search by mass and again display the first ten responses.

```{r, msready-mass}
methane_mass <- get_msready_by_mass(start = 16.0313,
                                    end = 16.0314,
                                    API_key = my_key)
length(methane_mass)
methane_mass[1:10]
```

We follow up with searching by DTXCID and display the first 10 responses.

```{r, msready-dtxcid}
methane_dtxcid <- get_msready_by_dtxcid(DTXCID = 'DTXCID705545',
                                        API_key = my_key)
length(methane_dtxcid)
methane_dtxcid[1:10]
```

Now, we show that the returned lists of DTXSIDs are in fact identical from all three calls to the API.

```{r, methane-check}
print(identical(methane_dtxcid, methane_formula))
print(identical(methane_dtxcid, methane_mass))
```

Observe that the lists of chemical DTXSIDs returned from the three functions are identical. One should note that when using the function `get_msready_by_mass`, the start and end range are sensitive to very small changes, so extra attention needs to be given to the input of this function.

#### Batch searching and combining functions

Before moving to the next set of functions, a demonstration of using more than one of the API functions may be helpful. With the previous set of functions, each returned a list of DTXSIDs. However, what these chemicals are is unclear without either memorizing vast sets of DTXSIDs or accessing information using this identifier. The latter is more tractable, so we demonstrate how to combine functions to achieve identification of the returned chemicals.

Here we combine the data.tables from the `get_chemical_details()` calls for a list of ten DTXSIDs.

```{r, methane-chemical-details-batch}
methane_df_list <- get_chemical_details_batch(DTXSID = methane_mass[1:10],
                                              API_key = my_key)
methane_df <- data.table::rbindlist(methane_df_list)
methane_df
```
If instead we just want the information from a single column, we can specify that as follows.

```{r, mol-formula-batch}
methane_formula_list <- sapply(methane_df_list, function(t) {
  t$molFormula
})
methane_formula_list
```
### Chemical List Resource

The next set of functions allow the user to access publicly available lists of chemicals. We first display the list of chemical lists and the search by the type of chemical lists available.

```{r, all-public-lists}
public_lists <- get_all_public_chemical_lists(API_key = my_key)
head(public_lists)
```

Next, we display a filtered set of lists focusing on federal lists.

```{r, federal-public-lists}
federal_public_lists <- get_chemical_lists_by_type(type = 'federal',
                                                   API_key = my_key)
head(federal_public_lists)
```

We can then access information about a specific list, for instance 'BIOSOLIDS2021'.

```{r, biosolids2021}
biosolids_2021_info <- get_public_chemical_list_by_name(list_name = 'BIOSOLIDS2021',
                                                        API_key = my_key)
biosolids_2021_info
```

And we can access the chemicals contained in the list as follows.

```{r, biosolids2021-chemicals}
biosolids_2021_chemicals <- get_chemicals_in_list(list_name = 'BIOSOLIDS2021',
                                                  API_key = my_key)
head(biosolids_2021_chemicals)
```

Finally, if one is interested in finding all lists that contain a particular chemical, one may do so by using the DTXSID of the chemical in question.

```{r, bpa-lists}
bpa_lists <- get_lists_containing_chemical(DTXSID = 'DTXSID7020182',
                                           API_key = my_key)
bpa_lists
```
### Chemical File Resource

Additional formats of chemical data are also available for interested users. These include Marvin files (.mrv), mol files (.mol), and image files.

We demonstrate this below.

We first grab the marvin file (.mrv) for BPA.

```{r, bpa-mrv}
bpa_mrv <- get_chemical_mrv(DTXSID = 'DTXSID7020182',
                            API_key = my_key)
if (requireNamespace("XML", quietly = TRUE)){
  XML::xmlTreeParse(bpa_mrv)
}
```

Next, we grab the mol file (.mol) for BPA.

```{r, bpa-mol}
bpa_mol <- get_chemical_mol(DTXSID = 'DTXSID7020182',
                            API_key = my_key)
cat(bpa_mol)
```

Finally, we grab an image file for BPA. Note that the image can be a .png or .svg format, and will default to png if none is specified.

```{r, bpa-image}
bpa_image_matrix <- get_chemical_image(DTXSID = 'DTXSID7020182',
                                       API_key = my_key)
if (requireNamespace("countcolors", quietly = TRUE)){
  countcolors::plotArrayAsImage(bpa_image_matrix)
}
```

If one wants to gather the synonyms of a given chemical, this is also avilable through the chemical APIs as demonstrated below.

```{r, synonym}
bpa_synonym <- get_chemical_synonym(DTXSID = 'DTXSID7020182',
                                    API_key = my_key)
head(bpa_synonym)
```


## Hazard APIs

Next, we explore the hazard APIs. TWe first demonstrate how to access human or ecological toxicity data for a given chemical. One can specify the chemical in question through the DTXSID.

### Hazard Resource

We first grab all such hazard data.

```{r, hazard}
bpa_hazard <- get_hazard_by_dtxsid(DTXSID = 'DTXSID7020182',
                                   API_key = my_key)
head(bpa_hazard)
```

Then we focus on human hazard data.

```{r, human-hazard}
bpa_human_hazard <- get_human_hazard_by_dtxsid(DTXSID = 'DTXSID7020182',
                                               API_key = my_key)
head(bpa_human_hazard)
```

And finally we focus on ecotox hazard data.

```{r, ecotox-hazard}
bpa_eco_hazard <- get_ecotox_hazard_by_dtxsid(DTXSID = 'DTXSID7020182',
                                              API_key = my_key)
head(bpa_eco_hazard)
```
### Skin and eye resource

Next, we demonstrate the skin and eye resource endpoint.

```{r, skin-eye}
bpa_skin_and_eye <- get_skin_eye_hazard(DTXSID = 'DTXSID7020182',
                                        API_key = my_key)
bpa_skin_and_eye
```

### Cancer summary

We then demonstrate the cancer summary endpoint.

```{r, cancer}
bpa_cancer <- get_cancer_hazard(DTXSID = 'DTXSID7020182',
                                API_key = my_key)
bpa_cancer
```

### Genetox resource

Finally we explore the genetox resource. This allos us to grab either summary or detail level genetox data. We first demonstrate the summary endpoint.

```{r, genetox-summary}
bpa_genetox_summary <- get_genetox_summary(DTXSID = 'DTXSID7020182',
                                           API_key = my_key)
bpa_genetox_summary
```

Next we demonstrate the details endpoint.

```{r, genetox-details}
bpa_genetox_details <- get_genetox_details(DTXSID = 'DTXSID7020182',
                                           API_key = my_key)
bpa_genetox_details
```



## Bioactivity APIs

The last set of APIs allow the user to access bioactivity information for a given chemical DTXSID or AEID.

### Bioactivity Resource

We first demonstrate obtaining bioactivity data based on an input DTXSID.

```{r, bioactivity-dtxsid}
bpa_bioactivity <- get_bioactivity_details(DTXSID = 'DTXSID7020182',
                                           API_key = my_key)
head(bpa_bioactivity)
```

If instead we want to search by assay id, we can do so using the same function.

```{r, bioactivity-aeid}
assay_id_search <- get_bioactivity_details(AEID = 42,
                                           API_key = my_key)
head(assay_id_search)
```


```{r, include=FALSE}
print(Sys.getenv('NOT_CRAN'))
```

